---
title: "Data Science Capstone - Movielense project"
author: "Julia Schröder"
date: "11/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = FALSE,
                     warning = FALSE,
                     message = FALSE,
                     root.dir = normalizePath("/Users/julia/sciebo/Kurse/2020-03_HarvardX_DataScience/HarvardX_DS/ml-10M100K"))

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(RColorBrewer)
library(kableExtra)

# presets
# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

# dl <- tempfile()
# download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
# 
# ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
#                  col.names = c("userId", "movieId", "rating", "timestamp"))
# 
# movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
# colnames(movies) <- c("movieId", "title", "genres")
# 
# # if using R 3.6 or earlier
# movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
#                                            title = as.character(title),
#                                            genres = as.character(genres))
# movielens <- left_join(ratings, movies, by = "movieId")
# 
# # Validation set will be 10% of MovieLens data
# set.seed(1)
# # if using R 3.5 or earlier, use `set.seed(1)` instead
# test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
# edx <- movielens[-test_index,]
# temp <- movielens[test_index,]
# 
# # Make sure userId and movieId in validation set are also in edx set
# validation <- temp %>% 
#   semi_join(edx, by = "movieId") %>%
#   semi_join(edx, by = "userId")
# 
# # Add rows removed from validation set back into edx set
# removed <- anti_join(temp, validation)
# edx <- rbind(edx, removed)
# 
# rm(dl, ratings, movies, test_index, temp, movielens, removed)
# 
# fwrite(edx, "ML_edx_training_data.tsv", sep = "\t")                      
# fwrite(validation, "ML_validation_data.tsv", sep = "\t")                
edx <- fread("ML_edx_training_data.tsv")
# validation <- fread("ML_validation_data.tsv")
```

# Introduction  
  
### Description of dataset  
The dataset used in this analysis is called "movielens" and comprises `r nrow(edx)` ratings of `r n_distinct(edx$movieId)` given by `r n_distinct(edx$userId)` users. These are the occurences per rating:  
```{r intro_overview, echo = FALSE}
edx %>%
  count(movieId, sort = TRUE) %>% 
  ggplot(aes(y = n)) +
  geom_histogram() 
  scale_x_log10()
  coord_flip()
# want to show that few movies have received a lot of ratings and many movies have received only few ratings (book 33.7.1)
  
edx %>%
  count(userId, sort = TRUE) %>% 
  ggplot(aes(y = n)) +
  geom_histogram() 
  scale_x_log10()
# want to show that fewer users have rated many movies and most users have only rated few movies

# show number of ratings per rating point
ggplot(data = edx, aes(x = as.factor(rating), fill = as.factor(rating))) +
  geom_bar() +
  scale_y_continuous(breaks = c(0, 5e5, 1e6, 1.5e6, 2e6, 2.5e6), 
                     labels = c("0", "0.5", "1", "1.5", "2", "2.5")) +
  theme_minimal() +
  scale_fill_viridis_d(guide = FALSE) +
  labs(x = "Rating", y = "Count in millions")
```
  
  
These are the 10 highest rated movies:  
```{r intro_top10, echo = FALSE}
edx %>%
  group_by(movieId) %>%
  summarize(rating_avg = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(rating_avg)) %>% 
  ungroup() %>% 
  inner_join(select(edx, movieId, title), by = "movieId") %>%
  distinct(movieId, .keep_all = TRUE) %>% 
  select(-movieId) %>%
  head(10) %>% 
  kable(format.args = list(big.mark = ",")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
```
  
  
These are the average ratings per genre:  
```{r intro_genre, echo = FALSE, warning=FALSE}
genres <- str_split(edx$genres, fixed("|"), simplify = TRUE) %>%
  as.data.frame() %>%
  bind_cols(select(edx, movieId)) %>% 
  distinct(.keep_all = TRUE) %>% 
  gather(col, genre, -movieId) %>% 
  filter(genre != "") %>% 
  select(-col)

genres_rating <- genres %>%
  left_join(select(edx, movieId, rating), by = "movieId") %>%
  group_by(genre) %>%
  summarize(rating_avg = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(rating_avg))
  # head(10) %>% 
  # kable(format.args = list(big.mark = ",", digits = 3)) %>% 
  # kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")

ggplot(data = genres_rating, aes(x = reorder(genre, -rating_avg, sum), y = rating_avg, fill = reorder(genre, -rating_avg, sum))) +
  geom_col() +
  theme_minimal() +
  scale_fill_viridis_d(guide = FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Genres", y = "Rating average per genre")
```  


### Summary of the project's goal

Goal of this project was to build a movie recommendation system. It will predict movie ratings by training a machine learning (ML) algorithm using the available ratings from the 10M version of the MovieLens dataset. Herefore, this dataset was divided into the training and test datasets and ...  
  
### Key analysis steps  
#### Data cleaning  
Data cleaning involved the following steps:  

* changing the timestamp to a date object  
* extracting the year of the movie release from the title  
* extracting the genres from the nested column  


#### Model selection  
First of all, the dataset is divided into training and test sets by using 90% for training and 10% for testing. Additionally, only users and movies that have appeared in the training data were included in the test dataset.

The simplest model would predict the same rating for all movies regardless of user. However, the type of movie and user are also playing a role and need to be accounted for in a model.


#### Model evaluation  


# Methods

explaination of process + techniques used: incl. data cleaning, data exploration + visualization
insights gained
modeling approach

### Data cleaning  
The timestamp was converted to a date object using the "as_datetime" function from the lubridate package.  
The release year and title of the movie was extracted from the title column using the "str_match" function from the stringr package.  
The genres were unnested by splitting the genres column by the "|" string and joining the data back using the movieId column.  
This is the head of the final dataframe used for training the ML algorithms:  
```{r methods_clean, echo = FALSE}
title_str <- as.data.frame(str_match(edx$title, pattern =  "(.*)\\s\\((\\d{4})\\)"))

edx_clean <- edx %>%
  mutate(rating_date = as_date(timestamp),
         movie_year = as.character(title_str[,3]),
         movie_title = as.character(title_str[,2])) %>% 
  left_join(genres, by = "movieId") %>%
  rename(movie_genre = genre) %>%
  select(userId, rating, rating_date, movieId, movie_title, movie_year, movie_genre)
rm(title_str)
head(edx_clean, 10) %>%
  kable(format.args = list(big.mark = ",")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
```
  
### Model training  
The dataset was split into training and test data by splitting 10% of the dataset for testing. Within the training data, cross validation was used to determine ...

The model was evaluated using the residual mean squared error:
$$
RMSE = \root{\frac{1/N}\sigma{u,i}(\hat{y}_{u,i}-<_{u,i})^2}
$$

```{r method_model, echo = FALSE, warning=FALSE}
edx_batch <- sample_frac(edx_clean, 0.01)
edx_batch <- edx_batch %>%
  group_by(movieId, rating_date) %>%
  nest(data = c(movie_genre)) %>%
  rename(movie_genre = data) %>%
  ungroup()
head(edx_batch)

test_index <- createDataPartition(edx_clean$rating, times=1, p=0.2, list=FALSE)
edx_test <- edx_clean[test_index,]
edx_train <- edx_clean[-test_index,]

edx_test <- edx_test %>%
  semi_join(edx_train, by = "movieId") %>%
  semi_join(edx_train, by = "userId")

```



# Results  

### Presentation of modeling results  

##### Basic model  
```{r results_mu, echo=FALSE}
mu <- mean(edx_train$rating)
rmse <- function(true_ratings, predicted_ratings){
    sqrt(mean((true_ratings - predicted_ratings)^2))
  }
rmse_mu <- rmse(edx_test$rating, mu)
rmse_all <- tibble(method = "mu_hat", RMSE = rmse_mu)
```  

The most basic model for predicting the movie ratings $Y_{u,i}$ would be the average of all ratings $\hat{\mu}$ and the independent sampmling errors $\epsilon_{u,i}$:

$$
Y_{u,i}=\mu+\epsilon_{u,i}
$$

where $\hat{\mu}$ is `r mu`. The RSME is `r rmse_mu`.

##### Movie effect  
```{r results_movie, echo=FALSE, warning=FALSE}
lambdas <- seq(0,10,0.25)

rmses_movie <- sapply(lambdas, function(l){
  b_movie <- edx_train %>%
    group_by(movieId) %>%
    summarize(b_movie = sum(rating - mu)/(n()+l))
  
  preds_movie <- edx_test %>%
    left_join(b_movie, by = "movieId") %>%
    mutate(pred = mu + b_movie) %>%
    .$pred
  rmse(edx_test$rating, preds_movie)
})

rmse_movie <- rmse(edx_test$rating, preds_movie)
rmse_all <- rbind(rmse_all, tibble(method = "+ movie effect", RMSE = rmse_movie))
```
Since certain movies are rated higher than others, the addition of a movie specific effect $b_i$ is added to the model: 

$$
Y_{u,i}=\mu+b_i+\epsilon_{u,i}
$$

The least squares estimate for this case is simply the average of the true rating $Y_{u,i}$ minus the average rating $\hat{\mu}$ per movie. This is the variation of these estimates:  
```{r results_movie_plot, echo=FALSE}
ggplot(b_movie, aes(x = b_movie, fill = b_movie)) +
  geom_histogram(bins = 10, color = "black", fill = "lightblue") + # change color!
  theme_minimal() 
```

Adding the movie effect ot the model improves the RMSE to `r rmse_movie`.  

##### User effect  
```{r results_user, echo=FALSE, warning=FALSE}
b_user <- edx_train %>%
  left_join(b_movie, by = "movieId") %>% 
  group_by(userId) %>%
  summarize(b_user = mean(rating - mu - b_movie))
preds_user <- edx_test %>%
  left_join(b_movie, by = "movieId") %>%
  left_join(b_user, by = "userId") %>% 
  mutate(pred = mu + b_movie + b_user) %>%
  .$pred
rmse_user <- rmse(edx_test$rating, preds_user) 
rmse_all <- rbind(rmse_all, tibble(method = "+ user effect", RMSE = rmse_user))

```

Similarily, certain users generally rate higher or lower. We can model this as the user effect $b_u$:
$$
Y_{u,i}=\mu+b_i+b_u+\epsilon_{u,i}
$$

This effect can be estimated by computing the average of each rating $Y_{u,i}$ minus the average rating $\hat{\mu}$ and minus the movie effect $b_i$ per user. This improves the RMSE further to `r rmse_user`.

##### Time effect  
smooth function of $d_{ui}$
```{r results_time, echo=FALSE, warning=FALSE}
ggplot(edx_batch, aes(x=rating_date, y = rating)) +
  geom_point() +
  geom_smooth()

b_time <- edx_train %>%
  left_join(b_movie, by = "movieId") %>% 
  left_join(b_user, by = "userId") %>% 
  group_by(rating_date) %>%
  summarize(b_time = mean(rating - mu - b_movie - b_user))

preds_time <- edx_test %>%
  left_join(b_movie, by = "movieId") %>%
  left_join(b_user, by = "userId") %>% 
  mutate(b_time = )
  group_by(movieId, userId) %>%
  mutate(pred = mu + b_movie + b_user + sum(b_genre)) %>%
  .$pred
rmse_genre <- rmse(edx_test$rating, preds_genre) 
rmse_all <- rbind(rmse_all, tibble(method = "+ genre effect", RMSE = rmse_user))

```

##### Genre effect  
sum of genre effects of all genres of the respective movie
```{r results_genre, echo=FALSE, warning=FALSE}
b_genre <- edx_train %>%
  left_join(b_movie, by = "movieId") %>% 
  left_join(b_user, by = "userId") %>% 
  group_by(movie_genre) %>%
  summarize(b_genre = mean(rating - mu - b_movie - b_user))
preds_genre <- edx_test %>%
  left_join(b_movie, by = "movieId") %>%
  left_join(b_user, by = "userId") %>% 
  left_join(b_genre, by = "movie_genre") %>% 
  group_by(movieId, userId) %>%
  mutate(pred = mu + b_movie + b_user + sum(b_genre)) %>%
  .$pred
rmse_genre <- rmse(edx_test$rating, preds_genre) 
rmse_all <- rbind(rmse_all, tibble(method = "+ genre effect", RMSE = rmse_genre))

```

##### Year of movie release effect  
```{r results_year, echo=FALSE, warning=FALSE}
b_year <- edx_train %>%
  left_join(b_movie, by = "movieId") %>% 
  left_join(b_user, by = "userId") %>% 
  left_join(b_genre, by = "movie_genre") %>% 
  group_by(movie_year) %>%
  summarize(b_year = mean(rating - mu - b_movie - b_user - b_genre))
preds_year <- edx_test %>%
  left_join(b_movie, by = "movieId") %>%
  left_join(b_user, by = "userId") %>% 
  left_join(b_genre, by = "movie_genre") %>% 
  left_join(b_year, by = "movie_year") %>% 
  group_by(movieId, userId) %>%
  mutate(pred = mu + b_movie + b_user + sum(b_genre) + b_year) %>%
  .$pred
rmse_year <- rmse(edx_test$rating, preds_year) 
rmse_all <- rbind(rmse_all, tibble(method = "+ year effect", RMSE = rmse_year))

```

### Discussion of model performance  
 RMSE < 0.86490


# Conclusion

### Brief summary of report  

### Limitations  

### Future work  